---
title: "Your own data: from a tree, occurrences, and climate rasters to ancestral niches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your own data: from a tree, occurrences, and climate rasters to ancestral niches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.2, dpi = 110
)
```

The round-trip vignette simulates everything. This one starts where most
users start: **you already have data** -- a phylogeny, a table of
occurrence points, and climate rasters -- and you want to reconstruct the
evolution of the niche over the clade.

The example is a real clade bundled with the package: six *Vulpes* foxes,
GBIF occurrences, an ERA5-Land bioclimatic window, and per-species
accessible-area polygons derived from ecoregions. The files are small and
live in `inst/extdata/vulpes/` -- and each one comes from a public
service, so the same recipe works for your own clade:

* **climate**: yearly ERA5-Land bioclim GeoTIFFs from
  `https://bioclim.ecoseek.org` (`/api/download/{year}/{var}_{year}.tif`),
  averaged over 1991--2020 -- the bundled `vulpes_climate.tif` is that
  product cropped and coarsened;
* **accessible areas**: `oneearthr::oe_layer("ecoregions")` (RESOLVE
  Ecoregions 2017) -- each species' $M_s$ is the dissolved union of the
  ecoregions containing its occurrences;
* **occurrences/tree**: GBIF + VertLife, frozen into the bundle;
* **prior ellipses**: `xnicher::optimize_niche` per species, wrapped by
  `niche_warmstart()`.

The runnable recipe is `data-raw/vulpes_bundle.R` in the source tree; the
retrieval steps are shown below where they belong in the workflow.

## What arfun consumes (and what it does not)

arfun does **not** read coordinates or rasters in the fitting functions.
The joint niche-evolution model consumes three things:

1. a `phylo` tree whose tip labels match your species names;
2. `env_occ_list`: a named list with one numeric matrix **per species** of
   environmental values *already extracted* at that species' occurrence
   points (rows = points, columns = variables);
3. `env_m_list`: the same, but for a sample of the **accessible
   environmental background** $M_s$ of each species.

Everything else -- reading files, intersecting points with rasters,
deciding what $M_s$ is -- is user-side work, and that is the honest
division of labour: those choices are scientific decisions, not plumbing.
This vignette walks through them.

## 1. Read the inputs

```{r inputs}
library(arfun)
library(ape)      # read.tree
library(terra)    # rast, vect, extract

extdata <- system.file("extdata/vulpes", package = "arfun")
# when working from the source tree:
# extdata <- "inst/extdata/vulpes"

tree   <- read.tree(file.path(extdata, "vulpes_tree.nwk"))
occ    <- read.csv(file.path(extdata, "vulpes_occurrences.csv"))
clim   <- rast(file.path(extdata, "vulpes_climate.tif"))   # bio01, bio12
m_poly <- vect(file.path(extdata, "vulpes_m_polygons.gpkg"))

tree
head(occ)
clim
```

Real files never align on the first try. The VertLife-derived tip labels
carry underscores; the occurrence table uses spaces:

```{r harmonize}
tree$tip.label <- gsub("_", " ", tree$tip.label)
stopifnot(all(occ$species %in% tree$tip.label))
plot(tree, main = "Vulpes clade"); axisPhylo()
```

## 2. Extract climate at the occurrence points

The bundled raster came from bioclim.ecoseek.org; for your own clade the
download is two lines per variable-year:

```{r ecoseek, eval = FALSE}
years <- 1991:2020
url <- function(v, y) sprintf(
  "https://bioclim.ecoseek.org/api/download/%d/%s_%d.tif", y, v, y)
stk <- rast(vapply(years, function(y) {
  f <- tempfile(fileext = ".tif")
  download.file(url("bio01", y), f, mode = "wb", quiet = TRUE); f
}, character(1)))
bio01_mean <- mean(stk)   # climatology: average the yearly fields
```

`terra::extract` pulls the raster values at each coordinate. arfun then
wants one matrix per species, in a named list, with names equal to the tip
labels:

```{r extract}
pts <- vect(occ, geom = c("lon", "lat"), crs = "EPSG:4326")
env_at_occ <- extract(clim, pts, ID = FALSE)

env_occ_list <- split(as.data.frame(env_at_occ), occ$species)
env_occ_list <- lapply(env_occ_list,
                       function(x) as.matrix(x[complete.cases(x), ]))
sapply(env_occ_list, nrow)
```

## 3. Standardize the environmental axes

bio01 arrives in tenths of kelvin and bio12 in mm of precipitation; the
model works on any units, but comparable scales make priors meaningful and
rates interpretable. Standardize with the **clade-pooled** occurrence
moments (so every species shares the same units), and keep the constants:
you need them again to interpret parameters or project back.

```{r scale}
all_occ <- do.call(rbind, env_occ_list)
mu0 <- colMeans(all_occ)
sd0 <- apply(all_occ, 2, sd)
env_occ_list <- lapply(env_occ_list, function(x) scale(x, mu0, sd0))
```

## 4. Declare M: the accessible background

This is the step the paper stresses: the likelihood normalizes each
species' niche against the environments it could plausibly have reached.
The bundled polygons declare $M_s$ as the dissolved union of RESOLVE
ecoregions containing that species' records, built with `oneearthr`
(shipped offline, no download needed at `scale = "small"`):

```{r oneearthr, eval = FALSE}
library(oneearthr)
eco <- oe_layer("ecoregions")                # or "bioregions"/"realms"
hit <- relate(eco, pts, "intersects")
m_of <- function(s)
  aggregate(eco[rowSums(hit[, occ$species == s, drop = FALSE]) > 0, ])
m_poly <- do.call(rbind, lapply(tree$tip.label, function(s) {
  m <- m_of(s); m$species <- s; m
}))
```

Choosing `oe_layer("realms")` or `"bioregions"` instead changes the
declared accessible domain -- deliberately coarser $M$ hypotheses, useful
as sensitivity analyses. Sample background points inside each polygon:

```{r background}
bg_list <- lapply(seq_len(nrow(m_poly)), function(i) {
  clim_m <- mask(crop(clim, m_poly[i]), m_poly[i])
  s <- spatSample(clim_m, size = 2000, method = "regular",
                  na.rm = TRUE, warn = FALSE)
  as.matrix(s[complete.cases(s), ])
})
names(bg_list) <- m_poly$species
bg_list <- lapply(bg_list, function(x) scale(x, mu0, sd0))
sapply(bg_list, nrow)
```

If all species share one accessible domain, draw a single sample and pass
the *identical matrix* to every species -- `shared_background_sample()`
exists for that. Independent subsamples of a shared $M$ create Monte Carlo
"holes" that leak likelihood into a spurious $\sigma \to 0$ mode.

## 5. Sparse species: `mask_species`

Presence-only data bound the niche from above; a species with a handful of
points carries essentially no shape information. Look at the counts:
`Vulpes pallida` has `r nrow(env_occ_list[["Vulpes pallida"]])` usable
point -- not enough to identify an ellipse. `mask_species` keeps it in the
phylogenetic prior but drops it from the likelihood, so its niche is
*predicted from its relatives* -- an identifiability limit turned into a
testable prediction.

```{r prepare}
sparse <- names(which(sapply(env_occ_list, nrow) < 10))
sparse

stan_data <- prepare_phylo_niche_data(
  env_occ_list, bg_list, tree, mask_species = sparse)
str(stan_data[c("S", "P", "N_occ_total", "N_m_total", "masked")])
```

## 6. Prior ellipses: per-species `xnicher` fits

Before the joint model, fit each species on its own. `niche_warmstart()`
runs `xnicher::optimize_niche()` (CRAN) per species against its declared
$M_s$, keeps the fits that avoid the degenerate $\sigma \to 0$ mode, and
assembles a starting vector for the joint parameterization -- unfitted or
masked species start at the phylogenetic conditional mean of the fitted
ones:

```{r warmstart}
warm <- niche_warmstart(stan_data)
warm$per_species
```

`per_species` is the interesting object on its own: the independent
M-restricted niche estimate of each species, i.e. what the data would say
*without* phylogenetic pooling. Comparing it with the joint fit below
shows what the evolutionary model adds -- and `fitted = FALSE` flags the
species whose ellipse is prior-driven rather than identified.

## 7. Fast screen: the MAP fit

Before paying for MCMC, run `niche_multistart()`: a space-filling design
over the joint posterior optimized with the compiled C++ objective, plus
the warm-start vector so the search begins at the per-species optima. It
takes seconds for a clade this size and flags boundary behaviour
(collapsed/sliver/inflated niches) via `niche_boundary_report()`:

```{r map}
ms <- niche_multistart(stan_data, n_starts = 16, seed = 1,
                       extra_starts = warm$theta)
ms$map_lp
niche_boundary_report(stan_data, ms$map)
```

Read the flags before reading the numbers. Here every fitted species
reports `ok` -- the declared $M_s$ contains enough accessible contrast to
identify the ellipses -- and only the masked `Vulpes pallida` is flagged.
Contrast this with what a tighter or misspecified $M$ produces (the
simulation vignette's `outside` cells): the optimizer then slides along a
flat ridge and the report flags `at_floor`/`sigma_huge`. Boundary flags
are a diagnostic of the $M$ hypothesis, not just of the optimizer. The
Bayesian fit below still matters: the MAP is a point on a ridge, and the
posterior carries the mass the point estimate discards.

## 8. Full posterior: `fit_evolution`

The Bayesian fit samples tip niches, ancestral states, and evolutionary
rates jointly. On a laptop this takes on the order of ten minutes for six
tips; on a cluster, pre-compile the model once and pass it via `mod =`.

```{r mcmc, eval = FALSE}
fit <- fit_evolution(stan_data, model = "bm_evolving",
                     chains = 2, iter_warmup = 1000,
                     iter_sampling = 1000, seed = 7,
                     mod = compile_model("bm_evolving"))
```

The vignette ships the result of exactly this call, run on KU HPC:

```{r cache}
cache_file <- file.path(extdata, "vulpes_bm_evolving_cache.rds")
cache <- if (file.exists(cache_file)) readRDS(cache_file) else NULL
if (!is.null(cache)) {
  dg <- cache$diagnostics
  cat("cached fit:", cache$args$model, "| draws:", nrow(cache$draws),
      "| divergences:", sum(dg$num_divergent),
      "| max-treedepth hits:", sum(dg$num_max_treedepth),
      "| min E-BFMI:", round(min(dg$ebfmi), 2), "\n")
}
```

Report the diagnostics, not just the estimates: on this dataset 5
transitions diverge and most hit the maximum treedepth -- the sampler
reaches the answers but moves slowly through the funnel-shaped posterior
(the phylogenetically pooled variance geometry). ESS stays above ~120 and
$\hat{R} \le 1.02$, so the marginals are usable, but a deeper treedepth
or a non-centered reparameterization is the honest next step for
production use.

The Stan model samples *tip* states; internal-node posteriors are
reconstructed post-hoc by pushing each draw through the BM conditional
formula -- $\mathbb{E}[y_{anc}\mid y_{tips}] = \mu_{anc} + W(y_{tips} -
\mu_{anc})$ with $W = V_{N,T}V_{T,T}^{-1}$ from the phylogenetic
covariance. The cache already contains these marginals
(`cache$node_posterior`); the computing recipe lives in
`?prepare_phylo_niche_data` and the package vignette sources.

```{r nodes, fig.height = 5}
if (!is.null(cache)) {
  np <- cache$node_posterior
  depth <- node.depth.edgelength(tree)
  par(mfrow = c(2, 3), mar = c(3.2, 3.2, 2, 0.6))
  panels <- list(list(np$mu1, expression(mu[1])),
                 list(np$mu2, expression(mu[2])),
                 list(np$ls1, expression(log~sigma[1])),
                 list(np$ls2, expression(log~sigma[2])),
                 list(np$zr,  expression(z[rho])))
  for (pn in panels) {
    m <- pn[[1]]$mean; s <- pn[[1]]$sd
    d <- depth[cache$node_ids]
    plot(d, m, pch = 19, cex = 0.7,
         xlab = "rootward depth", ylab = pn[[2]],
         ylim = range(m + 2 * s, m - 2 * s),
         main = paste(pn[[2]], ": internal nodes"))
    segments(d, m - 2 * s, d, m + 2 * s, col = "grey70")
    points(d, m, pch = 19, cex = 0.7)
  }
}
```

## 9. Back to geography

`project_single_niche()` evaluates the Mahalanobis suitability of any
state -- tip or reconstructed node -- on a raster. Project the root state
(un-standardizing back to raster units first, since `clim` is raw):

```{r project}
if (!is.null(cache)) {
  np <- cache$node_posterior
  root <- which.max(depth[cache$node_ids])
  mu_z  <- c(np$mu1$mean[root], np$mu2$mean[root])
  sg_z  <- c(exp(np$ls1$mean[root]), exp(np$ls2$mean[root]))
  rho   <- tanh(np$zr$mean[root])
  # un-z-score: x_raw = x_z * sd0 + mu0, so mu_raw = mu_z*sd0+mu0 and
  # Sigma_raw = D sd0 %*% Sigma_z %*% D sd0
  mu_raw <- mu_z * sd0 + mu0
  Sz <- diag(sg_z); R <- matrix(c(1, rho, rho, 1), 2)
  Sigma_raw <- diag(sd0) %*% Sz %*% R %*% Sz %*% diag(sd0)
  suit <- project_single_niche(mu_raw, Sigma_raw, clim)
  plot(suit, main = "reconstructed root niche, projected to geography")
}
```

## Honest scope

* The fitted ellipse is the niche **conditioned on the declared $M_s$** --
  change the background and the answer changes. The round-trip vignette
  shows what a flat background does to coverage.
* A suitability map of an ancestral state is a projection of a
  statistical estimate, not a paleodistribution reconstruction.
* The joint evolutionary models are currently two-dimensional
  (`bio01`, `bio12` here). `fit_niche()` (single species, no phylogeny)
  accepts more dimensions.
* Presence-only data bound the niche from above; sparse species belong in
  `mask_species`, not in the likelihood.
