---
title: "Round trip: evolving and reconstructing a fundamental niche"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Round trip: evolving and reconstructing a fundamental niche}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.2, dpi = 110
)
```

This vignette does the full round trip in the idealized case where the
**fundamental niche itself is observed**: a niche with five evolving
parameters -- centroid $\mu_1, \mu_2$, log tolerances
$\log\sigma_1, \log\sigma_2$, and correlation $z_\rho=\mathrm{atanh}(\rho)$
-- drifts by Brownian motion along a ten-species tree, generates
occurrences, and is then reconstructed from the occurrences alone,
*including the internal nodes*.  The point is to see, parameter by
parameter, what the evolutionary process looks like forward in time and
what the reconstruction recovers backward.

The honest-scope box first: occurrences here are drawn directly from the
fundamental response over a shared, fully accessible environmental
domain -- the oracle case of the paper's Level-1/2 separation.  The
harder case, where each species sees only its own accessible area $M_s$
and ignoring $M$ manufactures a spurious basal niche shift, is the
M-structured null benchmark (see `validation/mnull/` in the paper repo).

```{r setup}
library(arfun)
library(ape)
set.seed(7)
```

## 1. A ten-species tree

```{r tree}
tr <- rtree(10, tip.label = paste0("sp", 1:10))
tr$edge.length <- tr$edge.length / max(node.depth.edgelength(tr))
plot(tr, main = "simulated clade (unit height)"); axisPhylo()
S <- length(tr$tip.label); P <- 2L
```

## 2. Forward trip: the niche evolves

All five niche parameters evolve as independent Brownian motions from a
root state.  We traverse the tree edge by edge -- the same transition
rule `simulate_niche_evolution()` uses under `process = "bm"`,
$x_\text{child} \sim N(x_\text{parent},\; q\, t)$ -- but keep the value
at *every* node so the full trait history is visible, not just the tips.

```{r forward}
truth_anc <- list(mu = c(0, 0), ls = log(c(0.5, 0.5)), zr = 0)
rates     <- list(mu = c(0.5, 0.5), ls = c(0.4, 0.4), zr = 0.5)

evolve_bm <- function(tree, anc, rates) {
  edge <- tree$edge; el <- tree$edge.length
  n_tot <- length(tree$tip.label) + tree$Nnode
  val <- list(mu = matrix(NA_real_, n_tot, 2),
              ls = matrix(NA_real_, n_tot, 2),
              zr = matrix(NA_real_, n_tot, 1))
  val$mu[S + 1, ] <- anc$mu; val$ls[S + 1, ] <- anc$ls
  val$zr[S + 1, ] <- anc$zr
  for (e in order(edge[, 1])) {              # parents precede children
    pa <- edge[e, 1]; ch <- edge[e, 2]; t <- el[e]
    for (b in names(val))
      val[[b]][ch, ] <- val[[b]][pa, ] +
        rnorm(ncol(val[[b]]), 0, sqrt(rates[[b]] * t))
  }
  val
}

node_val <- evolve_bm(tr, truth_anc, rates)
depth <- node.depth.edgelength(tr)           # time from root, tips=1

# tip-level truth, natural scale
true_mu    <- node_val$mu[1:S, ]
true_sigma <- exp(node_val$ls[1:S, ])
true_rho   <- tanh(node_val$zr[1:S, ])
```

```{r traitgram-truth, fig.height=6}
par(mfrow = c(2, 3), mar = c(3.2, 3.2, 2.2, 0.6))
panels <- list(list(node_val$mu[, 1], expression(mu[1])),
               list(node_val$mu[, 2], expression(mu[2])),
               list(node_val$ls[, 1], expression(log~sigma[1])),
               list(node_val$ls[, 2], expression(log~sigma[2])),
               list(node_val$zr[, 1], expression(z[rho])))
for (pn in panels) {
  x <- pn[[1]]
  plot(depth, x, pch = ifelse(seq_along(x) <= S, 19, 21),
       col = "grey30", xlab = "time (root -> tip)", ylab = pn[[2]],
       main = paste("true BM trajectory:", pn[[2]]))
  for (e in seq_len(nrow(tr$edge)))
    segments(depth[tr$edge[e, 1]], x[tr$edge[e, 1]],
             depth[tr$edge[e, 2]], x[tr$edge[e, 2]], col = "grey60")
  points(depth, x, pch = ifelse(seq_along(x) <= S, 19, 21),
         col = "grey30")
}
```

Each line segment is one branch of the tree; dots are tips, open circles
internal nodes.  This is what "the fundamental niche evolves" concretely
means in the model: five Brownian paths sharing one topology.

## 3. Occurrences from the fundamental niche

The accessible domain is a dense environmental grid -- every species
sees the *same* domain, so there is no accessibility structure to
confound.  Occurrences are sampled proportional to the niche intensity
(Jimenez & Soberon weighting over the domain).

```{r occurrences}
g <- expand.grid(bio1 = seq(-4, 4, len = 60),
                 bio12 = seq(-4, 4, len = 60))
env_grid <- as.matrix(g)
occ_list <- lapply(seq_len(S), function(s)
  virtual_species_occurrences(env_grid, mu = true_mu[s, ],
                              sigma = true_sigma[s, ],
                              rho = true_rho[s], n_occ = 150))
names(occ_list) <- tr$tip.label

# shared background sample: the SAME rows for every species
bg <- env_grid[sample.int(nrow(env_grid), 2000), ]
bg_list <- lapply(tr$tip.label, function(s) bg)
names(bg_list) <- tr$tip.label
```

```{r ellipses-truth, fig.height=5}
niche_ellipse <- function(mu, sigma, rho, level = 0.68, n = 120) {
  Sig <- matrix(c(sigma[1]^2, rho * prod(sigma),
                  rho * prod(sigma), sigma[2]^2), 2, 2)
  L <- t(chol(Sig))
  th <- seq(0, 2 * pi, length.out = n)
  t(mu + sqrt(qchisq(level, 2)) * (L %*% rbind(cos(th), sin(th))))
}
plot(env_grid, pch = ".", col = "grey90",
     main = "true fundamental niches (68% ellipses)",
     xlab = "bio1", ylab = "bio12")
for (s in seq_len(S)) {
  points(occ_list[[s]], pch = ".", col = adjustcolor(s + 1, alpha.f = 0.4))
  lines(niche_ellipse(true_mu[s, ], true_sigma[s, ], true_rho[s]),
        col = s + 1, lwd = 2)
}
```

## 4. Backward trip: reconstruct from the tips alone

The joint MAP fit (`niche_multistart`, `noncentered_bounded`
parameterization) recovers the five-parameter BM model: per-species
niche parameters plus the evolutionary hyperparameters
($\mu_\text{anc}$, $\log\sigma_\text{anc}$, $z_{\rho,\text{anc}}$ and the
three rate blocks).  Decoding is the non-centered transform the Stan
model uses internally:
$\mu_{\cdot,k} = \mu_{\text{anc},k} + \sqrt{q_{\mu,k}}\, L_C z_{\mu,k}$.

```{r fit}
dat <- prepare_phylo_niche_data(occ_list, bg_list, tr)
fit <- niche_multistart(dat, "noncentered_bounded",
                        n_starts = 12, seed = 7, verbose = FALSE)
fit

pr  <- arfun:::.niche_unpack(fit$map, S, P, "bnd")
L_C <- t(chol(dat$C))
fit_mu <- sapply(1:P, function(k)
  pr$mu_anc[k] + sqrt(pr$rate_mu[k]) * (L_C %*% pr$mu_raw[, k]))
fit_ls <- sapply(1:P, function(k)
  pr$log_sigma_anc[k] + sqrt(pr$rate_ls[k]) * (L_C %*% pr$log_sigma_raw[, k]))
fit_sigma <- exp(fit_ls)
fit_zr <- as.numeric(pr$z_rho_anc +
                     sqrt(pr$rate_rho) * (L_C %*% pr$z_rho_raw))
fit_rho <- tanh(fit_zr)
```

## 5. The internal nodes: ancestral reconstruction

Under BM the reconstruction is a GLS conditional mean: for node $a$,
$\hat Y_a = \hat\mu_\text{anc} + C_{aT}C_{TT}^{-1}(Y_T -
\hat\mu_\text{anc}\mathbf{1})$, where $C$ is shared path length from the
root.  The predictor weights $C_{aT}C_{TT}^{-1}$ are **rate-free** --
BM ancestral reconstruction does not depend on the fitted rate, which is
why rate estimates and node reconstructions can be inspected
independently.

```{r nodes}
# full tip+node covariance: V[a,b] = depth of MRCA(a,b)
mrc  <- mrca(tr, full = TRUE)
V    <- matrix(depth[mrc], nrow = nrow(mrc))
C_TT <- V[1:S, 1:S]
recon_node <- function(tip_vals, anc_est, a) {
  w <- solve(C_TT, V[a, 1:S])
  anc_est + sum(w * (tip_vals - anc_est))
}
node_ids <- (S + 1):(S + tr$Nnode)
rec <- lapply(node_ids, function(a) c(
  mu  = sapply(1:P, function(k) recon_node(fit_mu[, k], pr$mu_anc[k], a)),
  ls  = sapply(1:P, function(k) recon_node(fit_ls[, k], pr$log_sigma_anc[k], a)),
  zr  = recon_node(fit_zr, pr$z_rho_anc, a)))
rec <- do.call(rbind, rec)

# joint conditional over the NON-ROOT internal nodes (needed for path
# draws):  mean_N = anc + W(Y_T - anc),  cov_N = q (C_NN - C_NT C_TT^-1 C_TN).
# The root is excluded: its "state" is the parameter mu_anc itself, so
# its conditional variance given the fit is zero and would make the
# joint covariance singular.  Its uncertainty is the posterior of
# mu_anc, handled by the Stan version below.
nr_ids <- node_ids[-1]
C_NT <- V[nr_ids, 1:S]; C_NN <- V[nr_ids, nr_ids]
C_TT_inv <- solve(C_TT)
W_NT <- C_NT %*% C_TT_inv
recon_joint <- function(tip_vals, anc_est, q) {
  mu <- anc_est + drop(W_NT %*% (tip_vals - anc_est))
  Sg <- q * (C_NN - W_NT %*% t(C_NT))
  list(mean = mu, cov = (Sg + t(Sg)) / 2)
}
```

```{r traitgram-compare, fig.height=7}
par(mfrow = c(2, 3), mar = c(3.2, 3.2, 2.2, 0.6))
fitted_all <- list(
  rbind(fit_mu[, 1], rec[, 1]), rbind(fit_mu[, 2], rec[, 2]),
  rbind(fit_ls[, 1], rec[, 3]), rbind(fit_ls[, 2], rec[, 4]),
  c(fit_zr, rec[, 5]))
for (i in seq_along(panels)) {
  x <- panels[[i]][[1]]
  plot(depth, x, pch = NA, xlab = "time (root -> tip)",
       ylab = panels[[i]][[2]],
       main = paste(panels[[i]][[2]], ": truth vs reconstruction"))
  for (e in seq_len(nrow(tr$edge)))
    segments(depth[tr$edge[e, 1]], x[tr$edge[e, 1]],
             depth[tr$edge[e, 2]], x[tr$edge[e, 2]], col = "grey75")
  points(depth[1:S], fitted_all[[i]][1:S], pch = 19,
         col = "firebrick", cex = 0.9)
  points(depth[node_ids], fitted_all[[i]][S + 1:tr$Nnode],
         pch = 21, bg = "white", col = "firebrick", cex = 0.9)
}
legend("topleft", c("true path", "reconstructed tip", "reconstructed node"),
       lty = c(1, NA, NA), pch = c(NA, 19, 21), col = c("grey60", "firebrick", "firebrick"),
       bty = "n", cex = 0.8)
```

### Why the reconstructed path does not look like the true path

The comparison above looks disappointing if read as "the point estimate
should trace the simulated path" -- but that is the wrong expectation.
The GLS estimator is a **conditional mean**: it is the smoothed
compromise that minimizes squared error, not a draw from the process.
A realized Brownian path is jagged; its conditional mean is not.  The
honest object is the **posterior over the node states**, which is
closed-form under BM: with a flat prior on the root (the GLS estimate
*is* the posterior mean),

$$\mathrm{Var}(Y_N \mid Y_T) =
q\,(C_{NN} - C_{NT}C_{TT}^{-1}C_{TN}).$$

So instead of one wrong-looking path we report the posterior ensemble:
draws from the joint conditional are genuine Brownian-bridge paths
through the fitted tips, and the true path should look like a plausible
member of that ensemble.  The checkable criterion is **coverage**, not
pointwise similarity.

```{r node-draws}
# posterior path draws for each parameter block
block_specs <- list(
  list(fit_mu[, 1], pr$mu_anc[1],        pr$rate_mu[1]),
  list(fit_mu[, 2], pr$mu_anc[2],        pr$rate_mu[2]),
  list(fit_ls[, 1], pr$log_sigma_anc[1], pr$rate_ls[1]),
  list(fit_ls[, 2], pr$log_sigma_anc[2], pr$rate_ls[2]),
  list(fit_zr,      pr$z_rho_anc,        pr$rate_rho))
set.seed(9); n_draw <- 40
node_draws <- lapply(block_specs, function(bs) {
  rj <- recon_joint(bs[[1]], bs[[2]], bs[[3]])
  L  <- t(chol(rj$cov))
  rj$mean + L %*% matrix(rnorm(length(nr_ids) * n_draw),
                         length(nr_ids), n_draw)
})
```

```{r traitgram-posterior, fig.height=7}
par(mfrow = c(2, 3), mar = c(3.2, 3.2, 2.2, 0.6))
for (i in seq_along(panels)) {
  x <- panels[[i]][[1]]
  plot(depth, x, pch = NA, xlab = "time (root -> tip)",
       ylab = panels[[i]][[2]],
       main = paste(panels[[i]][[2]], ": posterior paths vs truth"))
  for (d in seq_len(n_draw)) {                 # posterior path ensemble
    v <- fitted_all[[i]]                       # tips + mean path as base
    v[nr_ids] <- node_draws[[i]][, d]          # draw non-root nodes
    for (e in seq_len(nrow(tr$edge)))
      segments(depth[tr$edge[e, 1]], v[tr$edge[e, 1]],
               depth[tr$edge[e, 2]], v[tr$edge[e, 2]],
               col = adjustcolor("firebrick", alpha.f = 0.13))
  }
  for (e in seq_len(nrow(tr$edge)))            # posterior mean path
    segments(depth[tr$edge[e, 1]], fitted_all[[i]][tr$edge[e, 1]],
             depth[tr$edge[e, 2]], fitted_all[[i]][tr$edge[e, 2]],
             col = "firebrick", lwd = 1.2)
  for (e in seq_len(nrow(tr$edge)))            # true path on top
    segments(depth[tr$edge[e, 1]], x[tr$edge[e, 1]],
             depth[tr$edge[e, 2]], x[tr$edge[e, 2]],
             col = "grey25", lwd = 1.4)
  points(depth, x, pch = ifelse(seq_along(x) <= S, 19, 21),
         col = "grey25")
}
```

```{r coverage}
# 95% band coverage of the true NON-ROOT node values (the root's
# conditional sd is zero by construction -- its estimate is mu_anc,
# whose error we report directly)
true_nodes <- cbind(node_val$mu[node_ids, ], node_val$ls[node_ids, ],
                    node_val$zr[node_ids, ])
sd_nodes <- sapply(seq_along(block_specs), function(i)
  sqrt(diag(recon_joint(block_specs[[i]][[1]], block_specs[[i]][[2]],
                        block_specs[[i]][[3]])$cov)))
in_band <- abs(rec[-1, ] - true_nodes[-1, ]) <= 1.96 * sd_nodes
data.frame(parameter = c("mu1", "mu2", "ls1", "ls2", "zr"),
           node_sd = round(colMeans(sd_nodes), 3),
           coverage_95 = unname(colMeans(in_band)),
           row.names = NULL)
cat("root: estimate vs truth |",
    "|mu_anc - truth| =", round(abs(pr$mu_anc - truth_anc$mu), 3),
    " |ls_anc - truth| =", round(abs(pr$log_sigma_anc - truth_anc$ls), 3),
    " |zr_anc - truth| =", round(abs(pr$z_rho_anc - truth_anc$zr), 3), "\n")
```

Now the true paths (grey) sit inside the posterior path bundle (red) —
which is the meaningful notion of correspondence.  Two honest caveats:
the bands inherit the inflated fitted rate ($\hat q \approx 3\times$),
so they are *conservative* (nominal 95% covering at 100%); and they
condition on the fitted tip values, ignoring tip uncertainty.

The full Bayesian version runs the joint Stan model
(`niche_bm_evolving_sigma`) on this dataset and pushes each posterior
draw through the same conditional formula, marginalizing over
hyperparameter *and* tip uncertainty.  On KU HPC (job 29921878, 2
chains, 2000 draws, zero divergences) the marginal node bands come out
**narrower**, not wider: mean sd ≈ 0.19 vs ≈ 0.31, with coverage
0.89–1.00 across the eight non-root nodes — nearly nominal rather than
conservative.  The reason is instructive: the posterior rate
($\hat q \approx 0.2$–0.7, against a truth of 0.4–0.5) is far less
inflated than the MAP corner ($\hat q \approx 1.5$–1.9), so the full
posterior partially *cures* the rate inflation that made the analytic
bands over-wide.  Point estimates barely move (node RMSE 0.07–0.21 in
both) — what the posterior buys is **calibrated uncertainty**, not
better means.  Draws and analyzer live in `validation/bmrt/` of the
paper repository.

The levers that actually shrink node bands are structural, not
computational: denser tip sampling (shorter average tip-to-node
distance), more occurrences per species (tighter fitted tips), or
fossil/ancient data that puts observations at the nodes themselves.

## 6. Round-trip summary

```{r summary-table}
data.frame(
  parameter = c("mu_anc[1]", "mu_anc[2]", "ls_anc[1]", "ls_anc[2]",
                "zr_anc", "rate_mu1", "rate_mu2", "rate_ls1",
                "rate_ls2", "rate_rho"),
  truth = c(truth_anc$mu, truth_anc$ls, truth_anc$zr,
            rates$mu, rates$ls, rates$zr),
  MAP   = round(unname(c(pr$mu_anc, pr$log_sigma_anc, pr$z_rho_anc,
                  pr$rate_mu, pr$rate_ls, pr$rate_rho)), 3))
cat("tip mu RMSE   :", round(sqrt(mean((fit_mu - true_mu)^2)), 3), "\n")
cat("tip ls  RMSE  :", round(sqrt(mean((fit_ls - node_val$ls[1:S, ])^2)), 3), "\n")
cat("tip zr  RMSE  :", round(sqrt(mean((fit_zr - node_val$zr[1:S, ])^2)), 3), "\n")
node_rmse <- sqrt(rowMeans((rec - cbind(node_val$mu[node_ids, ],
                                       node_val$ls[node_ids, ],
                                       node_val$zr[node_ids, ]))^2))
cat("node RMSE (mu1 mu2 ls1 ls2 zr):\n"); print(round(node_rmse, 3))
```

```{r ellipse-root}
plot(env_grid, pch = ".", col = "grey95", main = "ancestral niche: truth vs MAP",
     xlab = "bio1", ylab = "bio12")
lines(niche_ellipse(truth_anc$mu, exp(truth_anc$ls), tanh(truth_anc$zr)),
      col = "grey30", lwd = 2)
lines(niche_ellipse(pr$mu_anc, exp(pr$log_sigma_anc), tanh(pr$z_rho_anc)),
      col = "firebrick", lwd = 2, lty = 2)
legend("topright", c("true root niche", "reconstructed root niche"),
       col = c("grey30", "firebrick"), lty = c(1, 2), lwd = 2, bty = "n")
```

## 7. Same trip under an OU process

BM wanders without bound.  The alternative the package implements is
Ornstein--Uhlenbeck: each parameter is pulled toward an optimum
$\theta$ with strength $\alpha$,

$$x_\text{child} \sim N\!\left(\theta +
(x_\text{parent}-\theta)e^{-\alpha t},\;
\frac{q}{2\alpha}\left(1-e^{-2\alpha t}\right)\right),$$

the same transition `simulate_niche_evolution()` uses under
`process = "ou"`.  We rerun the trip on the *same* tree with the same
rates, $\alpha = 2.5$ per block (half-life $\approx 0.28$ on a
unit-height tree — attraction is clearly visible), optimum
$\theta = (0.4, -0.3,\ \log 0.6, \log 0.4,\ 0.3)$ and a stationary
start (root value $=\theta$).

```{r ou-forward}
theta_ou <- list(mu = c(0.4, -0.3), ls = log(c(0.6, 0.4)), zr = 0.3)
alpha_ou <- list(mu = c(2.5, 2.5), ls = c(2.5, 2.5), zr = 2.5)

evolve_ou <- function(tree, anc, rates, alpha, theta) {
  edge <- tree$edge; el <- tree$edge.length
  n_tot <- length(tree$tip.label) + tree$Nnode
  val <- list(mu = matrix(NA_real_, n_tot, 2),
              ls = matrix(NA_real_, n_tot, 2),
              zr = matrix(NA_real_, n_tot, 1))
  val$mu[S + 1, ] <- anc$mu; val$ls[S + 1, ] <- anc$ls
  val$zr[S + 1, ] <- anc$zr
  for (e in order(edge[, 1])) {
    pa <- edge[e, 1]; ch <- edge[e, 2]; t <- el[e]
    for (b in names(val)) {
      a <- alpha[[b]]
      m <- theta[[b]] + (val[[b]][pa, ] - theta[[b]]) * exp(-a * t)
      v <- rates[[b]] / (2 * a) * (-expm1(-2 * a * t))
      val[[b]][ch, ] <- m + rnorm(ncol(val[[b]]), 0, sqrt(v))
    }
  }
  val
}

set.seed(73)
node_val_ou <- evolve_ou(tr, theta_ou, rates, alpha_ou, theta_ou)
true_mu_ou    <- node_val_ou$mu[1:S, ]
true_sigma_ou <- exp(node_val_ou$ls[1:S, ])
true_rho_ou   <- tanh(node_val_ou$zr[1:S, ])
```

```{r traitgram-ou-truth, fig.height=6}
theta_flat <- c(theta_ou$mu, theta_ou$ls, theta_ou$zr)
par(mfrow = c(2, 3), mar = c(3.2, 3.2, 2.2, 0.6))
panels_ou <- list(list(node_val_ou$mu[, 1], expression(mu[1])),
                  list(node_val_ou$mu[, 2], expression(mu[2])),
                  list(node_val_ou$ls[, 1], expression(log~sigma[1])),
                  list(node_val_ou$ls[, 2], expression(log~sigma[2])),
                  list(node_val_ou$zr[, 1], expression(z[rho])))
for (i in seq_along(panels_ou)) {
  x <- panels_ou[[i]][[1]]
  plot(depth, x, pch = NA, xlab = "time (root -> tip)",
       ylab = panels_ou[[i]][[2]],
       main = paste("true OU trajectory:", panels_ou[[i]][[2]]))
  abline(h = theta_flat[i], lty = 2, col = "steelblue", lwd = 1.5)
  for (e in seq_len(nrow(tr$edge)))
    segments(depth[tr$edge[e, 1]], x[tr$edge[e, 1]],
             depth[tr$edge[e, 2]], x[tr$edge[e, 2]], col = "grey60")
  points(depth, x, pch = ifelse(seq_along(x) <= S, 19, 21),
         col = "grey30")
}
```

Compare with the BM traitgram above: instead of a widening fan, OU
paths oscillate around the dashed optimum — the "hedgehog" signature.
Tip spread is bounded by the stationary variance $q/2\alpha$, not by
tree height.

### The misspecified trip: fitting BM to OU data

Now generate occurrences exactly as before and fit — deliberately — the
**BM** joint MAP (the only process the local optimizer implements; the
OU model is a Stan/CmdStan fit).  This is the paper's failure mode F3,
process misspecification, in miniature.

```{r ou-occurrences-fit}
occ_ou <- lapply(seq_len(S), function(s)
  virtual_species_occurrences(env_grid, mu = true_mu_ou[s, ],
                              sigma = true_sigma_ou[s, ],
                              rho = true_rho_ou[s], n_occ = 150))
names(occ_ou) <- tr$tip.label

dat_ou <- prepare_phylo_niche_data(occ_ou, bg_list, tr)
fit_ou <- niche_multistart(dat_ou, "noncentered_bounded",
                           n_starts = 8, seed = 7, verbose = FALSE)
fit_ou

pr_ou <- arfun:::.niche_unpack(fit_ou$map, S, P, "bnd")
fmu_ou <- sapply(1:P, function(k)
  pr_ou$mu_anc[k] + sqrt(pr_ou$rate_mu[k]) * (L_C %*% pr_ou$mu_raw[, k]))
fls_ou <- sapply(1:P, function(k)
  pr_ou$log_sigma_anc[k] + sqrt(pr_ou$rate_ls[k]) *
    (L_C %*% pr_ou$log_sigma_raw[, k]))
fzr_ou <- as.numeric(pr_ou$z_rho_anc +
                     sqrt(pr_ou$rate_rho) * (L_C %*% pr_ou$z_rho_raw))
```

```{r traitgram-ou-compare, fig.height=7}
rec_ou <- lapply(node_ids, function(a) c(
  mu = sapply(1:P, function(k) recon_node(fmu_ou[, k], pr_ou$mu_anc[k], a)),
  ls = sapply(1:P, function(k) recon_node(fls_ou[, k], pr_ou$log_sigma_anc[k], a)),
  zr = recon_node(fzr_ou, pr_ou$z_rho_anc, a)))
rec_ou <- do.call(rbind, rec_ou)

fitted_ou <- list(
  rbind(fmu_ou[, 1], rec_ou[, 1]), rbind(fmu_ou[, 2], rec_ou[, 2]),
  rbind(fls_ou[, 1], rec_ou[, 3]), rbind(fls_ou[, 2], rec_ou[, 4]),
  c(fzr_ou, rec_ou[, 5]))
par(mfrow = c(2, 3), mar = c(3.2, 3.2, 2.2, 0.6))
for (i in seq_along(panels_ou)) {
  x <- panels_ou[[i]][[1]]
  plot(depth, x, pch = NA, xlab = "time (root -> tip)",
       ylab = panels_ou[[i]][[2]],
       main = paste(panels_ou[[i]][[2]], ": OU truth vs BM fit"))
  abline(h = theta_flat[i], lty = 2, col = "steelblue", lwd = 1.5)
  for (e in seq_len(nrow(tr$edge)))
    segments(depth[tr$edge[e, 1]], x[tr$edge[e, 1]],
             depth[tr$edge[e, 2]], x[tr$edge[e, 2]], col = "grey75")
  points(depth[1:S], fitted_ou[[i]][1:S], pch = 19,
         col = "firebrick", cex = 0.9)
  points(depth[node_ids], fitted_ou[[i]][S + 1:tr$Nnode],
         pch = 21, bg = "white", col = "firebrick", cex = 0.9)
}
```

```{r ou-summary}
data.frame(
  parameter = c("mu_anc[1]", "mu_anc[2]", "ls_anc[1]", "ls_anc[2]",
                "zr_anc", "rate_mu1", "rate_mu2", "rate_ls1",
                "rate_ls2", "rate_rho"),
  theta_or_truth = c(theta_ou$mu, theta_ou$ls, theta_ou$zr,
                     rates$mu, rates$ls, rates$zr),
  BM_MAP_on_OU = round(unname(c(pr_ou$mu_anc, pr_ou$log_sigma_anc,
                                pr_ou$z_rho_anc, pr_ou$rate_mu,
                                pr_ou$rate_ls, pr_ou$rate_rho)), 3))
cat("tip mu RMSE (OU data, BM fit):",
    round(sqrt(mean((fmu_ou - true_mu_ou)^2)), 3), "\n")
```

Three things to read off the misspecified fit:

- **Tip values still land correctly** — the occurrence clouds pin them
  down regardless of the process prior; tip RMSE stays small.
- **The root estimate lands near θ anyway**: under stationary OU the
  tips cluster around the optimum, so the BM weighted-mean
  reconstruction lands close to θ — a *correct point estimate for the
  wrong reason*.  What it cannot express is the regression-to-θ
  geometry of the true paths.
- **The rates are uninterpretable**: the BM layer reads both the
  bounded stationary spread *and* per-species estimation noise as
  diffusion, inflating `rate_*` roughly threefold — the same
  noise-soaking mechanism as the correctly specified BM run — and there
  is no $\alpha$ slot where the attraction could be reported.  The
  fitted "process" is fiction: OU's pull is silently rewritten as a
  featureless diffusion.

The correct-fit path is the OU Stan model:
`fit_evolution(stan_data, model = "ou_evolving")` fits the joint
occurrence + OU likelihood with $\alpha$ and $\sigma^2$ as free
parameters (the `mu_anc` parameter doubles as the optimum θ).  For this
exact dataset that model was run on KU HPC (CmdStan 2.36, SLURM job
29921840, 2 chains, 1000 warmup + 1000 sampling, 2 divergences, all
$\hat R = 1.00$).  Posterior medians for the attraction strength:
$\hat\alpha_\mu = (1.96, 1.99)$, $\hat\alpha_{\log\sigma} = (1.47,
3.02)$, $\hat\alpha_\rho = 2.47$ — all against a truth of 2.5, with
wide 90% intervals (roughly 0.2–6.6): $\alpha$ is weakly identified
with ten tips but the posterior is centred correctly and, unlike the BM
fit, the parameter *exists to be estimated*.  The variance parameters
$\sigma^2$ come out inflated (medians $\hat\sigma^2_\mu = (1.31,
0.26)$, $\hat\sigma^2_{\log\sigma} = (0.48, 0.55)$,
$\hat\sigma^2_\rho = 0.71$ against truths of 0.25, 0.16, 0.25) — the
same noise-absorption mechanism as the BM rates.  Draws, diagnostics and the
analyzer live in `validation/ou/` of the paper repository.  The point
of the contrast: the right process *estimates* the attraction, the
wrong process hides it inside an inflated rate — which is why the paper
insists on reporting the fitted process alongside any ancestral
reconstruction.

## What to look at

- **Centroids ($\mu$)** are the best-identified parameters: occurrence
  clouds locate them directly, and node reconstructions track the true
  path closely.
- **Log tolerances and $z_\rho$** are noisier at the tips (a presence
  cloud pins down an ellipse's shape weakly) but the BM prior pools
  information across the tree -- the phylogeny is doing measurable work.
- **Rates** ($q_\mu$, $q_{ls}$, $q_\rho$) are the hardest: with ten tips
  the likelihood carries little information about variance parameters,
  so MAP rates sit inside wide priors.  In this run they come out
  inflated roughly threefold -- the raw deviations soak up per-species
  estimation noise, which the BM layer then reads as rate.  Treat them
  as order-of-magnitude statements.
- The residual caveats of the pipeline apply: this is a MAP fit, not a
  posterior; $\nu$ is the oracle uniform-over-domain case; and $P=2$.

