---
title: "Virtual species in a real bioregion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Virtual species in a real bioregion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5, dpi = 110
)
```

This vignette closes the simulation loop: **simulate a known
niche-evolution process inside a real accessible area, then recover it**.
It is the fastest way to learn what the joint phylo-niche model can and
cannot identify before analysing empirical clades.

The workflow:

1. Take a real bioregion polygon as the shared accessible area $M$
   (One Earth Bioregions Framework via `oneearthr`).
2. Build a plausible two-dimensional environmental space over $M$.
3. Evolve a virtual clade's niche under BM, OU and the
   independent-species null.
4. Draw occurrences with `virtual_species_occurrences()` and rebuild the
   posterior modes with `niche_multistart()`.
5. Compare recovered niches against the truth — including which species
   land on identifiability boundaries.

```{r setup}
library(arfun)
library(ape)
set.seed(42)
```

## 1. A real accessible area

We use the Andean Mountain Forests & Valleys bioregion (NT11) — home to
the hummingbird clades that motivated this machinery. If `oneearthr` or
`terra` are not installed we fall back to a synthetic rectangular $M$ so
the vignette still builds.

```{r bioregion}
have_spatial <- requireNamespace("oneearthr", quietly = TRUE) &&
                requireNamespace("terra", quietly = TRUE)

if (have_spatial) {
  bio <- oneearthr::oe_layer("bioregions")
  nt11 <- bio[bio$bioregions == "NT11", ]
  # rasterize the polygon footprint and keep cell centres as M points
  r <- terra::rast(nt11, resolution = 0.25)
  pts <- terra::as.points(terra::mask(r, nt11))
  coords <- terra::crds(pts)
  plot(nt11, main = "NT11 -- Andean Mountain Forests & Valleys")
  points(coords, pch = ".", col = "steelblue")
} else {
  coords <- expand.grid(lon = seq(-78, -66, 0.25),
                        lat = seq(-20, 6, 0.25))
  coords <- as.matrix(coords)
}
nrow(coords)
```

## 2. A plausible environmental space over $M$

Real clade pipelines extract ERA5/WorldClim values at each cell. Here we
synthesise two axes as smooth functions of geography -- an
elevation-flavoured temperature axis and a precipitation axis with
longitudinal structure -- then standardise. What matters is that $M$ is
*real*: its holes, corridor structure and extent are the bioregion's own.

```{r env}
lon <- scale(coords[, 1])[, 1]
lat <- scale(coords[, 2])[, 1]
temp <- -lat + 0.4 * lon + rnorm(length(lon), 0, 0.25)
prec <- 0.8 * lon - 0.3 * lat + rnorm(length(lon), 0, 0.35)
env_M <- scale(cbind(temp, prec))
colnames(env_M) <- c("bio1", "bio12")

plot(env_M, pch = ".", col = "grey70",
     main = "Environmental space of M")
```

## 3. Evolve a virtual clade

Six virtual species evolving under each process. `simulate_niche_evolution()`
returns the true per-species $(\mu, \sigma, \rho)$ in exactly the
parameterization the joint models estimate.

```{r simulate}
tree <- rtree(6, tip.label = paste0("vsp", 1:6))
tree$edge.length <- tree$edge.length / max(ape::vcv(tree))  # unit height

truths <- lapply(c("bm", "ou", "null"), function(pr)
  simulate_niche_evolution(tree, process = pr,
                           mu_anc = c(0, 0),
                           log_sigma_anc = log(c(0.45, 0.45)),
                           rate_mu = c(0.35, 0.35),
                           rate_log_sigma = c(0.3, 0.3),
                           rate_rho = 0.4,
                           alpha = c(2, 2, 2), seed = 42))
names(truths) <- c("bm", "ou", "null")

# draw occurrences: 120 per species, shared accessible area
occ_list <- lapply(truths, function(tr)
  lapply(seq_along(tree$tip.label), function(s)
    virtual_species_occurrences(env_M, mu = tr$mu[s, ],
                                sigma = tr$sigma[s, ], rho = tr$rho[s],
                                n_occ = 120)))
for (pr in names(occ_list)) names(occ_list[[pr]]) <- tree$tip.label

# the SAME background rows for every species -- never independent draws
bg_list <- lapply(tree$tip.label, function(s) env_M)
names(bg_list) <- tree$tip.label
```

```{r plot-truth, fig.show='hold'}
# 68% ellipse of a bivariate normal niche, no extra dependency
niche_ellipse <- function(mu, sigma, rho, level = 0.68, n = 100) {
  Sig <- matrix(c(sigma[1]^2, rho * prod(sigma),
                  rho * prod(sigma), sigma[2]^2), 2, 2)
  L <- t(chol(Sig))
  th <- seq(0, 2 * pi, length.out = n)
  t(mu + sqrt(qchisq(level, 2)) * (L %*% rbind(cos(th), sin(th))))
}

tr <- truths$bm
plot(env_M, pch = ".", col = "grey85",
     main = "BM niches (true)", xlab = "bio1", ylab = "bio12")
for (s in 1:6) {
  lines(niche_ellipse(tr$mu[s, ], tr$sigma[s, ], tr$rho[s]),
        col = s + 1, lwd = 2)
}
```

## 4. Recover: multistart MAP

```{r recover}
fits <- lapply(names(occ_list), function(pr) {
  d <- prepare_phylo_niche_data(occ_list[[pr]], bg_list, tree)
  niche_multistart(d, "noncentered_bounded", n_starts = 12,
                   seed = 7, verbose = FALSE)
})
names(fits) <- names(occ_list)
fits$bm
```

## 5. Truth vs recovery

```{r compare}
recover_traits <- function(fit) {
  # unpack the MAP back to (mu, sigma, rho) per species
  data <- prepare_phylo_niche_data(occ_list$bm, bg_list, tree)
  rep <- fit$boundary  # per-species sigma1, sigma2, rho at the MAP
  rep[, c("species", "sigma1", "sigma2", "rho")]
}

for (pr in c("bm", "ou", "null")) {
  tr <- truths[[pr]]; rec <- fits[[pr]]$boundary
  cat("\n--- process:", pr, "---\n")
  print(data.frame(
    species = rec$species,
    sigma_true = round(apply(tr$sigma, 1, min), 3),
    sigma_map  = round(pmin(rec$sigma1, rec$sigma2), 3),
    rho_true   = round(tr$rho, 2),
    rho_map    = round(rec$rho, 2),
    flag       = rec$flag))
}
```

## 6. Masking a virtual microendemic

Species whose occurrences occupy only a handful of environmental cells
cannot identify a 2-D ellipse -- the presence-only likelihood is
unbounded as $\sigma \to 0$. The recommended treatment is **masking**:
keep the species in the phylogenetic prior but drop its likelihood, so
the model *predicts* its niche from its relatives. The observed cluster
then becomes a held-out validation target.

```{r mask}
# crush vsp1 down to a microendemic: resample only 3 unique cells
occ_m <- occ_list$bm
mu_m <- truths$bm$mu[1, ]
d2 <- colSums((t(env_M) - mu_m)^2)
few <- order(d2)[1:3]
occ_m$vsp1 <- env_M[few, , drop = FALSE]

d_mask <- prepare_phylo_niche_data(occ_m, bg_list, tree,
                                   mask_species = "vsp1")
fit_mask <- niche_multistart(d_mask, "noncentered_bounded",
                             n_starts = 12, seed = 7, verbose = FALSE)
fit_mask$boundary
```

`vsp1` carries flag `"masked"`: no likelihood contribution, but the
phylogenetic prior still predicts a finite niche -- compare its
`sigma1/sigma2` against the true value used to simulate it.

## Reading the boundary flags

| flag | meaning |
|---|---|
| `ok` | interior optimum |
| `masked` | excluded from the likelihood, predicted by the prior |
| `at_floor` | $\sigma$ pinned at `sigma_floor` -- check `n_cells` before calling it microendemic |
| `rho_at_cap` | sliver-like ellipse, $\|\rho\| \to$ cap |
| `sigma_huge` | nearly uniform niche over $M$ |
| `few_cells` | narrow niche with < 15 unique cells -- candidate for masking |

The empirical pipeline adds one more diagnostic -- whether the *union* of
the shared background covers each species' cluster -- to separate real
microendemism from background-sampling holes (see
`?prepare_phylo_niche_data` and `?shared_background_sample`).
